---
title: "Texas Wide Open"
author: "Marvin Bernardino"
date: "2023-11-26"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

In 2013, Gov. Rick Perry launched the "Texas Wide Open For Business" campaign. Advertising campaigns were launched in major cities across California to lure them into moving to Texas (Texas Monthly, 2013).

In 2023, some Texans and other national figures are now blaming both California and these new migrants for whatever local problems these migrations may have caused.

Using some publicly available data, we will now try to answer whether Texas' inferiority complex is warranted or not.

## Data

The IRS publishes a yearly Population Migration Data. Every year, as taxes are filed, the IRS takes notes on address changes individuals report on their own income tax returns. 

The data normally includes:

- number of returns filed (approximates the number of households)
- number of personal exemptions claimed (approximates the number of invididuals)
- total adjusted gross income

The data is usually prepared year to year, meaning, it will take a lot of work to combine these data into bigger data sets, if year-to-year changes are to be analyzed.

Conveniently, user Patrick O'Connor in Kaggle, published a data set containing data from 1992 to 2020. 

## Kaggle Data by O'Connor

The data contains 9 fields, but some of these are just codes and abbreviations. The important fields include:

- Y1 (the year when the individual filed their last tax return)

- Y1_STATE_NAME (name of the state where the individual filed their last return)

- Y2 (the year when the individual filed their current tax return)

- Y2_STATE_NAME (name of the state the filer currently reside in)

- NUM_RETURNS (number of tax returns, approximates household)

- NUM_EXEMPTIONS (number of exemptions claimed, approximates number of individuals)

- AGI (adjusted gross income)

## Problems /  Objectives

The data will now able to show migration patterns, the number of households and number of individuals migrating to any state, and their household income.

We can now ask:

- How many individuals are migrating to Texas from California?

- What is the household income of these migrants?

- Was Gov. Rick Perry's advertising campaign effective?

- 
What other information can be gleamed from these data?

## Raw Data

```{r, echo=FALSE, message=FALSE}

library(tidyverse)
library(ggplot2)

irs_SOI_state_migraion_data <- read.csv("irs_SOI_state_migraion_data.csv")
glimpse(irs_SOI_state_migraion_data)
```

## California to Texas Migration Data, 1992 to 2020


```{r, echo=FALSE}
#filter California to Texas Data
CATXdata <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" & Y2_STATE_ABBR == "TX")

CATXdata <- CATXdata %>% select(Year = Y1, Californians = NUM_EXEMPTIONS, AGI, NUM_RETURNS)

#CATXdata <- CATXdata %>% mutate_at(c('Year'), as.factor)
options(scipen = 999)

#plot of California to Texas Data

ggplot(CATXdata, aes(Year, Californians)) + geom_point() + geom_line() + geom_vline(xintercept = 2013, linetype="dotted", color = "blue") + annotate(geom = "text", x = 2013, y = 100000, label = "2013") + theme_bw()
```

See the drop in the number of migrants moving to Texas after 2013.

## Texas vs. California Migration Data, 1992 to 2020

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%')

#filter Texas to California Data, Y1
TXCAdata <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "TX" & Y2_STATE_ABBR == "CA")

TXCAdata <- TXCAdata %>% select(Year = Y1, Texans = NUM_EXEMPTIONS, AGI, NUM_RETURNS)

#take out only CA to TX data for Graph 2
G2_CAdata <- CATXdata %>% select(Year, Californians)

#take out only TX to CA data for graph 2
G2_TXdata <- TXCAdata %>% select(Year, Texans)

#combine G2_CA to G2_TX dor graphing
G2_Data <- inner_join(G2_CAdata, G2_TXdata, by = join_by(Year))

options(scipen = 999)

G2_Data <- G2_Data %>% mutate_at(c('Year'), as.factor)


#plot of Texans leaving California with G1 at background

ggplot(G2_Data, aes(Year, Texans, group = 2)) + geom_line(color = "firebrick") + geom_line(aes(y=Californians), color = "dodgerblue3") + annotate(geom = "text", x = 8, y = 60000, label = "Californians Moving to Texas", color = "dodgerblue3") +  annotate(geom = "text", x = 5, y = 38000, label = "Texans Moving to California", color = "firebrick") + labs(y = "Number of Migrants") + theme_bw(base_size = 8)

```


## Top 3 Destinations for California Migrants, 1992 to 2020

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%')

TOP3 <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" & Y2_STATE_ABBR != "CA") %>% group_by(Y1) %>%  top_n(n=3, NUM_EXEMPTIONS) %>% select(State = Y2_STATE_NAME, NUM_EXEMPTIONS)

TOP3 <- TOP3 %>% mutate_at(c('Y1'), as.factor)

ggplot(TOP3, aes(x = forcats::fct_rev(Y1), NUM_EXEMPTIONS, fill = State)) + geom_col(position = "dodge") + coord_flip() + labs(y = "Number of Migrants", x = "Year") + theme_bw()

#ggplot(TOP3, aes(x=Y1, y=NUM_EXEMPTIONS, fill = Y2_STATE_NAME)) + geom_area(aes(fill=Y2_STATE_NAME, group = Y2_STATE_NAME))

```
Californians began moving to Texas in greater numbers in 2004.

## Percent Population of Texas and Eventual Destination
```{r}
allTEXANS <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "TX") %>% select(Y1, State = Y2_STATE_NAME, NUM_EXEMPTIONS)

allTEXANS <- allTEXANS %>% group_by(Y1) %>% mutate(TotalTX = sum(NUM_EXEMPTIONS))

allTEXANS <- allTEXANS %>% mutate_at(c('Y1'), as.factor)

allTEXANS <- allTEXANS %>% group_by(Y1) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalTX ) * 100) %>% top_n(n=5, Percent_Migration)

ggplot(allTEXANS, aes(x = Y1, y = Percent_Migration, fill = State)) + geom_col() + coord_cartesian(ylim = c(97,100))  + theme_bw(base_size = 8) + labs(y= "Percent Population", x = "Year")
```
Less than 3% of Texans move out of state. Note that some Texans are migrating overseas.


## Percent Population Californians and Eventual Destination
```{r}
allCA <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA") %>% select(Y1, State = Y2_STATE_NAME, NUM_EXEMPTIONS)


allCA <- allCA %>% group_by(Y1) %>% mutate(TotalCA = sum(NUM_EXEMPTIONS))

options(scipen = 999)

allCA <- allCA %>% mutate_at(c('Y1'), as.factor)

allCA <- allCA %>% group_by(Y1) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalCA ) * 100) %>% top_n(n=5, Percent_Migration)

stateCA_levels <- c("Texas", "Arizona", "Foreign", "Nevada", "Washington", "Oregon", "California")

#use position stack reverse as temporary solution
ggplot(allCA, aes(x = Y1, y = Percent_Migration, fill = State, order = Percent_Migration, group = State)) + geom_col( position = position_stack(reverse = TRUE)) +  theme_bw(base_size = 8) + labs(y= "Percent Population", x = "Year") + coord_cartesian(ylim = c(97, 100)) 

```
Less than 3% of Californians move out of state, same as with Texans.

## Household Income of Californians Migrating to Texas Part 1

Both the IRS and O'Connor's Kaggle Data provide AGI, the aggregate gross income of all transplants in thousands. If we divide this by the number of tax returns filed, we can now obtain the average household income in that particular year.

Average Household Income = (AGI * 1000 )/ NUM_RETURNS

## Household Income of Californians Migrating to Texas Part 2

```{r}
n1_CATXdata <- CATXdata %>% mutate_at(c('Year'), as.factor)

n1_CATXdata <- n1_CATXdata %>% mutate(Average_Household_Income = (AGI*1000)/(NUM_RETURNS))

ggplot(n1_CATXdata, aes(Year, y = Average_Household_Income, group = 3)) + geom_line(color = "dodgerblue3") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + labs(y = "Household Income in Dollars") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3")
```
After 2009, the average household income of California transplants start to increase.

## Applying Inflation Correction (in 2022 dollars) Part 1

To perform an inflation correction, we now obtain the average CPI data every year from 1992 to 2020 (FRED, 2023).

As an example calculation, we get the inflation rate in 2020 compared to 2020 dollars by:

2020 Inflation Rate = [(CPI_2022 - CPI_2020 )/ (CPI_2020)] * 100%

2020 Income in 2022 dollars = [2020 Income * (1+Inflation Rate) ]

We now apply this to the whole dataset.



## Applying Inflation Correction (in 2022 dollars) Part 2
```{r}
CPI = c(138.158, 142.058, 145.642, 149.758, 154.142, 157.558, 159.658, 163.233, 168.892,
        173.517, 175.875, 179.808, 184.483, 190.975, 197.142, 202.767, 211.053, 209.630, 213.967,
        221.575, 226.229, 229.324, 232.771, 231.810, 234.076, 239.051, 245.146, 249.222, 252.248)
#from https://www.ssa.gov/oact/STATS/avgcpi.html

new_AHIdata <- n1_CATXdata %>% select(Year, Average_Household_Income)

#add CPI to new_AHIdata
new_AHIdata <- cbind(new_AHIdata, CPI) 

#inflation rate in 2020 compared to 2022 = 
# [(CPI for 2022 - CPI for 2020 ) / (CPI for 2020)] x 100%

new_AHIdata <- new_AHIdata %>% mutate(new_AHIdata, inflation2022factor = ((287.984 - CPI)/CPI))

#income in 2020 (compared to 2022 dollars) = 
#(income in 2020) / (inflation rate in 2020 compared to 2022)

new_AHIdata <- new_AHIdata %>% mutate(Adjusted_in_2022dollars = Average_Household_Income*(1+inflation2022factor))

#head(new_AHIdata)

ggplot(new_AHIdata, aes(Year, y = Average_Household_Income, group = 3)) + geom_line(color = "dodgerblue3") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + geom_line(aes(y=Adjusted_in_2022dollars), color = "goldenrod3") + annotate(geom = "text", x = 8, y = 105000, label = "Average Household Income in 2022 dollars", color = "goldenrod3") + labs(y = "Household Income in Dollars") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3")
```

## Comparing California Median Household Income versus Texas-bound Californians

```{r}
#Median Household Income (2022 dollars, I believe!) from https://fred.stlouisfed.org/series/MEHOINUSCAA646N
MHI = c(34900, 34070, 35330, 37010, 38810, 39690, 40930, 43630, 46820, 47260, 47440, 49300, 49220, 51760, 55320, 55730, 57010, 56130, 54280, 53370, 57020, 60790, 60490, 63640, 66640, 70040, 70490, 78110, 77650)

new_AHIdata <- cbind(new_AHIdata, MHI)
```


```{r}
ggplot(new_AHIdata, aes(Year, y = Average_Household_Income, group = 4)) + geom_line(color = "dodgerblue3") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + geom_line(aes(y=Adjusted_in_2022dollars), color = "goldenrod3") + geom_line(aes(y=MHI), color = "firebrick4") + annotate(geom = "text", x = 20, y = 35000, label = "California Median Household Income in 2022 dollars", color = "firebrick4") +  annotate(geom = "text", x = 8, y = 105000, label = "Average Household Income in 2022 dollars", color = "goldenrod3") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3") + labs(y = "Household Income in Dollars")

```
Since 2009, the average household income of Californians moving to Texas have been increasing dramatically, after adjusting for inflation.


## Summary

Californians have been moving to Texas in greater numbers since 2004. At least 50,000 Californians are moving yearly since then. In 2020, 100,000 Californians have moved to Texas.

Between 2-3% of residents of both California and Texas move somewhere else.

The average California transplant to Texas is well-off, earning double the average household income in California.

In 2020, the average California transplant's household income has risen to 150,000.

Gov. Rick Perry's advertising campaign seems to have been triggered by a dip in migration from California, however, it is still true that Californians have been steadily moving to Texas on their own since 2004.


## References

Herzog, M., Texas Monthly Staff. (2013, February 5). *Rick Perry to Californians: Come Check Out Texas.* Texas Monthly.  https://www.texasmonthly.com/news-politics/rick-perry-to-californians-come-check-out-texas/

O'Connor, P. (2023, September). *IRS Migration Data - 1992 to 2020.* kaggle. https://www.kaggle.com/datasets/wumanandpat/irs-migration-data-1992-to-2020

Bureau of Labor Statistics. (2023, November). *Average CPI by Quarter and Year*. Social Security Administration. https://www.ssa.gov/oact/STATS/avgcpi.html

Federal Reserve Economic Data, St. Louis FED. (2023, November). *Median Household Income in California*. Economic Research, Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/MEHOINUSCAA646N