---
title: "Data 2401 Final Project"
author: "Marvin Bernardino"
date: "2023-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Texas Wide Open
Data 2401 Final Project

## Introduction

In 2013, Gov. Rick Perry launched the "Texas Wide Open For Business" campaign. Advertising campaigns were launched in major cities across California to lure them into moving to Texas (Texas Monthly, 2013).

In 2023, some Texans and other national figures are now blaming both California and these new migrants for whatever local problems these migrations may have caused.

Using some publicly available data, we will now try to answer whether these concerns are overblown or not.

## Data

The IRS publishes a yearly U. S. Population Migration Data (IRS, 2023). Every year, as taxes are filed, the IRS takes notes on address changes that tax filers report on their own income tax returns. 

The data normally includes:

- number of returns filed (approximates the number of households)
- number of personal exemptions claimed (approximates the number of invididuals)
- total adjusted gross income

The data is usually prepared per year, which means it will take a lot of work to combine these data into bigger data sets, if year-to-year changes are to be analyzed.

Conveniently, user Patrick O'Connor in Kaggle, published a data set containing data from 1992 to 2020. 

## Kaggle Data by Patrick O'Connor

Patrick O'Connor's dataset, uploaded in Kaggle, is a collection of IRS Migration Data from 1992 to 2020. The IRS dataset, however, starts in 1990 and also includes the counties from which the migrants moved from and their destination. The IRS dataset is massive but is only collected and reported on a yearly basis. Patrick O'Connor combined these datasets from 1993 to 2020 but removed the county codes. 

Note that the IRS only begun including AGI data in 1993. Hence, there will be no AGI figures to be reported before that year (IRS, 2023). Essentially, the data begins in 1993, after the tax filers completed their move.

Patrick O'Connor treats the year before as the start of the migration. However, this report will treat the next year as the starting year, after the tax filers completed the move (signaled by the filing of taxes with the new address).

The O'Connor dataset contains 9 fields, most of whom are state codes and abbreviations. The important fields include:

- **Y1** (the year when the individual filed their last tax return)

- **Y1_STATE_NAME** (name of the state where the individual filed their last return)

- **Y2** (the year when the individual filed their current tax return)

- **Y2_STATE_NAME** (name of the state the filer currently reside in)

- **NUM_RETURNS** (number of tax returns, approximates household)

- **NUM_EXEMPTIONS** (number of exemptions claimed, approximates number of individuals)

- **AGI** (adjusted gross income, the AGI represents the total income of all individuals who moved that particular year)

## Problems /  Objectives

The dataset shows migration patterns year to year, by tracking Y1 and Y2 changes. As a bonus, it also includes the approximated number of households, the approximated number of individuals migrating to any state, and their aprroximated household income.

These are some of the questions this report maybe ready to answer:

* What is the nature of Texas-California migration that can be gleamed from the data?

  + How many individuals are migrating to Texas from California?

  + Is Texas the number one destination for California migrants?

  + What is the net migration between Texas and California?

  + Do Californians make up the majority of those moving to Texas?
  
  + Are the residents of either state more likely to move to than the other?

* From which states are the majority of those moving to Texas?

* Using the AGI data, what other information can be gleamed from the migration data?
  
  + What is the household income of these migrants?
  
  + Does the migrant population differ from the population that stayed?
  
  + Does the migrant population differ from the residents of the state they are moving to?

* Was Gov. Rick Perry's advertising campaign effective?

* What other information can be gleamed from these data?

## Packages Used

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
```


## Raw Data

```{r, echo=FALSE}

irs_SOI_state_migraion_data <- read.csv("irs_SOI_state_migraion_data.csv")
glimpse(irs_SOI_state_migraion_data)
```

The O'Connor Kaggle Data is still arranged year to year, with each state, starting at **Y1_STATE...**, with an accompanying **Y2_STATE....** Note that **Y1_STATE** and **Y2_STATE** can be equal, which means that these are the tax filers who did not move to another state. 

**FIPS** codes stand for Federal Information Processing System Codes for States and Counties. The number uniquely identify geographic areas, for example, the state-level FIPS code for Texas is 48 (Federal Communications Commission). The O'Connor Dataset retained the state-level FIPS code but excluded the county-level FIPS codes. Hence, this report is currently limited to only analyzing state-level migration.

The **AGI** stands for aggregate gross income, and is reported in the thousands. This means that to obtain the true AGI in dollars, they have to be multiplied by 1000.

Each line in the data includes Y1 (previous year), Y2 (current year), Y1_State_x (previous state), Y2_State_x (current state), NUM_RETURNS (number of tax returns filed in Y2), AGI (the total income of all the filers in NUM_RETURNS).

## California to Texas Migration Data, 1993 to 2020

To obtain migration data from California to Texas, we use Y1_STATE_ABBR = "CA" and Y2_STATE_ABBR == "TX" to filter the data.


```{r, echo=FALSE}
#filter California to Texas Data
CATXdata <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" & Y2_STATE_ABBR == "TX")

CATXdata <- CATXdata %>% select(Year = Y2, Californians = NUM_EXEMPTIONS, AGI, NUM_RETURNS)

CATXdata
```
# Graph of California Migration to Texas

```{r echo=FALSE}
#CATXdata <- CATXdata %>% mutate_at(c('Year'), as.factor)
options(scipen = 999)

#plot of California to Texas Data

ggplot(CATXdata, aes(Year, Californians)) + geom_point() + geom_line() + geom_vline(xintercept = 2013, linetype="dotted", color = "blue") + annotate(geom = "text", x = 2013, y = 100000, label = "2013") + theme_bw()

```

At least 40,000 Californians are moving to Texas, yearly. In 2021, more than 100,000 Californians have moved to Texas. Note that there is a drop in migration around 2015, but Californians have been migrating steadily in increasing numbers since 2004.

## Texas vs. California Migration Data, 1993 to 2021

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%')

#filter Texas to California Data, Y1
TXCAdata <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "TX" & Y2_STATE_ABBR == "CA")

TXCAdata <- TXCAdata %>% select(Year = Y2, Texans = NUM_EXEMPTIONS, AGI, NUM_RETURNS)

#take out only CA to TX data for Graph 2
G2_CAdata <- CATXdata %>% select(Year, Californians)

#take out only TX to CA data for graph 2
G2_TXdata <- TXCAdata %>% select(Year, Texans)

#combine G2_CA to G2_TX dor graphing
G2_Data <- inner_join(G2_CAdata, G2_TXdata, by = join_by(Year))

options(scipen = 999)

G2_Data <- G2_Data %>% mutate_at(c('Year'), as.factor)


#plot of Texans leaving California with G1 at background

ggplot(G2_Data, aes(Year, Texans, group = 2)) + geom_line(color = "firebrick") + geom_line(aes(y=Californians), color = "dodgerblue3") + annotate(geom = "text", x = 8, y = 60000, label = "Californians Moving to Texas", color = "dodgerblue3") +  annotate(geom = "text", x = 5, y = 38000, label = "Texans Moving to California", color = "firebrick") + labs(y = "Number of Migrants") + theme_bw(base_size = 8)

```

Note that net migration between California and Texas favors California migration to Texas.

## Top 3 Destinations for California Migrants, 1993 to 2020.

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '100%')

TOP3 <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" & Y2_STATE_ABBR != "CA") %>% group_by(Y2) %>%  top_n(n=3, NUM_EXEMPTIONS) %>% select(State = Y2_STATE_NAME, NUM_EXEMPTIONS)

TOP3 <- TOP3 %>% mutate_at(c('Y2'), as.factor)

ggplot(TOP3, aes(x = forcats::fct_rev(Y2), NUM_EXEMPTIONS, fill = State)) + geom_col(position = "dodge") + coord_flip() + labs(y = "Number of Migrants", x = "Year") + theme_bw()

```
Note that Californians began moving to Texas in greater numbers starting in 2004.


## Top Destinations of Texas Migrants
```{r}
#without texans
sansTEXANS <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "TX" & Y2_STATE_ABBR != "TX") %>% select(Y2, State = Y2_STATE_NAME, NUM_EXEMPTIONS)

sansTEXANS <- sansTEXANS %>% group_by(Y2) %>% mutate(TotalnotTX = sum(NUM_EXEMPTIONS))


sansTEXANS <- sansTEXANS %>% group_by(Y2) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalnotTX ) ) %>% top_n(n=5, Percent_Migration)

ggplot(sansTEXANS, aes(x = Y2, y = Percent_Migration, fill = State)) + geom_col() + scale_y_continuous(labels = scales::percent) + facet_wrap(.~State) + labs(y= "Percent Migrants", x = "Year")

```
Note that a significant population of Texans leave for a foreign counties, from 1990s to 2010. Also note that, in 2007, there is a spike in Texans moving to Louisiana.


## Non-migrant Population of Texans

```{r}
allTEXANS <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "TX") %>% select(Y2, State = Y2_STATE_NAME, NUM_EXEMPTIONS)

allTEXANS <- allTEXANS %>% group_by(Y2) %>% mutate(TotalTX = sum(NUM_EXEMPTIONS))

allTEXANS <- allTEXANS %>% mutate_at(c('Y2'), as.factor)

allTEXANS <- allTEXANS %>% group_by(Y2) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalTX ) * 100) %>% top_n(n=1, Percent_Migration)

allTEXANS %>% select(Year = Y2, Percent_Migration)
  
ggplot(allTEXANS, aes(x = Y2, y = Percent_Migration, group = State, fill = State)) + geom_col()  + theme_bw(base_size = 8) + labs(y= "Percent Population", x = "Year") + geom_hline(yintercept = 97, linetype="solid", color = "blue") + annotate(geom = "text", x = 10, y = 97.2, label = "97%", color = "blue") +
  theme_bw(base_size = 7)
```
Between 2-3% of Texans migrate annually for other states.

## Top Destinations for California Migrants

```{r}
#without Californians
sansCAL <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" & Y2_STATE_ABBR != "CA") %>% select(Y2, State = Y2_STATE_NAME, NUM_EXEMPTIONS)

sansCAL <- sansCAL %>% group_by(Y2) %>% mutate(TotalnotCA = sum(NUM_EXEMPTIONS))


sansCAL <- sansCAL %>% group_by(Y2) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalnotCA ) ) %>% top_n(n=5, Percent_Migration)

ggplot(sansCAL, aes(x = Y2, y = Percent_Migration, fill = State)) + geom_col() + scale_y_continuous(labels = scales::percent) + facet_wrap(.~State) + labs(y= "Percent Migrants", x = "Year")
```
Texas features prominently as the top destination for California migrants, with at least 9% migrating annually. Foreign migration is also a feature, but to a lesser degree than Texas.

## Non-migrant Population of California

```{r}
allCA <- irs_SOI_state_migraion_data %>% filter(Y1_STATE_ABBR == "CA" ) %>% select(Y2, State = Y2_STATE_NAME, NUM_EXEMPTIONS)

allCA <- allCA %>% group_by(Y2) %>% mutate(TotalCA = sum(NUM_EXEMPTIONS))

allCA <- allCA %>% mutate_at(c('Y2'), as.factor)

allCA <- allCA %>% group_by(Y2) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / TotalCA ) * 100) %>% top_n(n=1, Percent_Migration)

allCA %>% select(Year = Y2, Percent_Migration)
  
ggplot(allCA, aes(x = Y2, y = Percent_Migration, group = State, fill = State)) + geom_col()  + theme_bw(base_size = 8) + labs(y= "Percent Population", x = "Year") + geom_hline(yintercept = 97, linetype="solid", color = "blue") + annotate(geom = "text", x = 10, y = 97.2, label = "97%", color = "blue")

```

Both California and Texas have similar levels of migration -- 97-98% do not migrate.


## Top State Migrants Moving to Texas

```{r}
TXmovers <- irs_SOI_state_migraion_data %>% filter(Y2_STATE_ABBR == "TX" & Y1_STATE_ABBR != "TX") %>% select(Y2, State = Y1_STATE_NAME, NUM_EXEMPTIONS)

TXmovers <- TXmovers %>% group_by(Y2) %>% mutate(Totalmovers = sum(NUM_EXEMPTIONS))


TXmovers <- TXmovers %>% group_by(Y2) %>% mutate(Percent_Migration = (NUM_EXEMPTIONS / Totalmovers ) ) %>% top_n(n=9, Percent_Migration)

ggplot(TXmovers, aes(x = Y2, y = Percent_Migration, fill = State)) + geom_col() + scale_y_continuous(labels = scales::percent) + facet_wrap(.~State) + labs(y= "Percent Migrants", x = "Year")

```
Californians feature prominently in this data. However, notice the jump in Louisiana in the mid-2000's,

## Louisiana Migrants moving to Texas, A Closer Look

```{r}
LA2006 <- TXmovers %>% filter(State == "LOUISIANA")



ggplot(LA2006, aes(x = Y2, y= NUM_EXEMPTIONS, fill = State)) + geom_col(fill='tan', color = 'black') + 
  labs(y= "Number of Migrants", x = "Year") +
  geom_vline(xintercept = 2006, linetype="longdash", color = 'black' ) + 
  annotate(geom = "text", x = 2008, y = 70000, label = "2006", color = 'black' ) 

  
```
Hurricane Katrina made landfall in New Orleans, Louisiana on August 29, 2005 (National Weather Service, 2022). Notice that there was also significant migration out of Texas back to Louisiana starting in 2007.

## Household Income of Californians Migrating To Texas

Both the IRS and O'Connor's Kaggle Data provide AGI, the aggregate gross income of all transplants expressed in thousands of dollars. If AGI is divided by the number of tax returns filed, we can now obtain the average household income for that particular year.

Average Household Income = (AGI * 1000 ) / NUM_RETURNS

```{r}
n1_CATXdata <- CATXdata %>% mutate_at(c('Year'), as.factor)

n1_CATXdata <- n1_CATXdata %>% mutate(Average_Household_Income = (AGI*1000)/(NUM_RETURNS))

ggplot(n1_CATXdata, aes(Year, y = Average_Household_Income, group = 3)) + geom_line(color = "dodgerblue3") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + labs(y = "Household Income in Dollars") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3")
```
After 2010, the average household income of California migrants start to increase.This is the reported data in that particular year and may not reflect inflation. Therefore, the data must be corrected for inflation.

## Applying Inflation Correction (in 2022 dollars) Part 1

To perform an inflation correction, the average CPI data every year from 1993 to 2020 was obtained from the Federal Reserve Data (FRED, 2023).

To perform an example calculation, the inflation rate in 2020 compared to 2022 dollars by:

2020 Inflation Rate = [(CPI_2022 - CPI_2020 )/ (CPI_2020)] * 100%

2020 Income in 2022 dollars = [2020 Income * (1 + Inflation Rate) ]

This can now be applied to the whole dataset.

```{r}
CPI = c(138.158, 142.058, 145.642, 149.758, 154.142, 157.558, 159.658, 163.233, 168.892,
        173.517, 175.875, 179.808, 184.483, 190.975, 197.142, 202.767, 211.053, 209.630, 213.967,
        221.575, 226.229, 229.324, 232.771, 231.810, 234.076, 239.051, 245.146, 249.222, 252.248)

#from https://www.ssa.gov/oact/STATS/avgcpi.html

new_AHIdata <- n1_CATXdata %>% select(Year, Average_Household_Income)

#add CPI to new_AHIdata
new_AHIdata <- cbind(new_AHIdata, CPI) 

#inflation rate in 2020 compared to 2022 = 
# [(CPI for 2022 - CPI for 2020 ) / (CPI for 2020)] x 100%

new_AHIdata <- new_AHIdata %>% mutate(new_AHIdata, inflation2022factor = ((287.984 - CPI)/CPI))

#income in 2020 (compared to 2022 dollars) = 
#(income in 2020) / (inflation rate in 2020 compared to 2022)

new_AHIdata <- new_AHIdata %>% mutate(Adjusted_in_2022dollars = Average_Household_Income*(1+inflation2022factor))

#head(new_AHIdata)

ggplot(new_AHIdata, aes(Year, y = Average_Household_Income, group = 3)) + geom_line(color = "dodgerblue3") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + geom_line(aes(y=Adjusted_in_2022dollars), color = "goldenrod3") + annotate(geom = "text", x = 8, y = 105000, label = "Average Household Income in 2022 dollars", color = "goldenrod3") + labs(y = "Household Income in Dollars") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3")
```
Applying inflation correction, the average household income for California migrants to Texas has been somewhat steady, averaging 80,000 dollars until 2009 -- when it starts to increase at a rapid rate.

## Comparing California Median Household Income versus Texas-bound Californians

To further explore the data, the average household income of the California migrant can be compared with the median household income of California. The St. Louis Fed provides the Median Household Income in California in "Current Dollars, Not Seasonally Adjusted," (FRED, 2023). Taking into consideration that the data is in "current dollars"  (2022), it is believed that the data has already been corrected for inflation. 

```{r}
#Median Household Income (2022 dollars, I believe!) from https://fred.stlouisfed.org/series/MEHOINUSCAA646N
MHI = c(34900, 34070, 35330, 37010, 38810, 39690, 40930, 43630, 46820, 47260, 47440, 49300, 49220, 51760, 55320, 55730, 57010, 56130, 54280, 53370, 57020, 60790, 60490, 63640, 66640, 70040, 70490, 78110, 77650)

new_AHIdata <- cbind(new_AHIdata, MHI)
```


```{r}
ggplot(new_AHIdata, aes(Year, y = Average_Household_Income, group = 4)) + geom_line(color = "dodgerblue3", linetype = "longdash") + theme_bw(base_size = 8) + labs(y = "Average Household Income") + geom_line(aes(y=Adjusted_in_2022dollars), color = "goldenrod3") + geom_line(aes(y=MHI), color = "firebrick4") + annotate(geom = "text", x = 20, y = 35000, label = "California Median Household Income in 2022 dollars", color = "firebrick4") +  annotate(geom = "text", x = 8, y = 105000, label = "Average Household Income in 2022 dollars", color = "goldenrod3") +  annotate(geom = "text", x = 12, y = 65000, label = "Average Household Income", color = "dodgerblue3") + labs(y = "Household Income in Dollars")


```
It can be expected that migrants as a whole -- should be better off than the average family. Moving itself, is an expensive venture, and can only be undertaken by those who can afford it. There is reason to believe that the California Median Household Income provided by the St. Louis Fed, has already been corrected for inflation.

During the latter part of 2010's, the average household income of the California migrants starts to increase rapidly, reaching nearly 160,000 dollars by 2021.

Note that the average household income of these California migrants dropped in the year 2001 which might signal that they have been significantly affected by the bursting of the Dot-com bubble. Also, note that the median California household's income was not affected as much by the Dot-com crisis. 

## Texas Median Household Income versus the California migrants

Using the same data from the St. Louis Fed, the average household income of the California migrant to Texas can be compared to the median household income of Texans.

```{r}
#Texas Median Household Income https://fred.stlouisfed.org/series/MEHOINUSTXA646N
MHITX = c(27950,28730, 30760, 32040, 33070, 35080, 35780, 38690, 38610,
          40860, 40150, 39270, 41400, 41420, 43310, 46050, 46490, 47480, 47270,
          49050, 51930, 51410, 53880, 56470, 58150, 60090, 59790, 67440, 68400)

new_AHIdata <- cbind(new_AHIdata, MHITX)
```

```{r}
ggplot(new_AHIdata, aes(Year, y = Average_Household_Income, group = 4)) + theme_bw(base_size = 8) + labs(y = "Average Household Income") + 
  geom_line(aes(y=Adjusted_in_2022dollars), color = "goldenrod3") + 
  geom_line(aes(y=MHITX), color = "firebrick4") + 
  annotate(geom = "text", x = 20, y = 35000, label = "Texas Median Household Income in 2022 dollars", color = "firebrick4") +  
  annotate(geom = "text", x = 12, y = 115000, label = "Average Household Income in 2022 dollars (California Transplant)", color = "goldenrod3") +  
  labs(y = "Household Income in Dollars")
```
The gap between the Texas median household income and the average household income of the California migrant is roughly double. After 2020, there is a almost a 100,000 income gap between the two groups.

## Summary

Californians have been moving to Texas in greater numbers since 2004. At least 50,000 Californians are moving yearly since then. In 2020, 100,000 Californians have started moving to Texas. These Californians filed their Texas in 2021 as full-fledged Texans.

Between 2-3% of residents of both California and Texas move somewhere else. If investigated further, the SOI data will reveal if this average is true for all states but that it is not within the scope of this report.

The data also shows that the average California transplant is incredibly well-off, earning double the median household income in California and Texas. In 2021, the average California migrant's household income has risen above 150,000 dollars.

With simple visualizations, some prominent real life events does make a mark in the dataset. How Hurricane Katrina affected the migration between Texas and Louisiana is very apparent in the data. The dip in the household income among California transplants during the year 2000 may indicate that the migrants are somehow connected to the tech industry (or they earn a lot of tech stocks). 

Announced in 2013, Gov. Rick Perry's advertising campaign was coincidentally succeeded by a dip in migration in 2014. Being that the data was collected and reported after taxes have been filed, meaning that there is a actually a dip in migration coinciding with the campaign. However, it is still true that Californians have been steadily moving to Texas their own since 2004.

## References

Bureau of Labor Statistics. (2023, November). *Average CPI by Quarter and Year*. Social Security Administration. https://www.ssa.gov/oact/STATS/avgcpi.html

Federal Communications Commision. (2023). *States and Counties FIPS codes*. https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt

Federal Reserve Economic Data, St. Louis FED. (2023, November). *Median Household Income in California*. Economic Research, Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/MEHOINUSCAA646N

Federal Reserve Economic Data, St. Louis FED. (2023, November). *Median Household Income in Texas*. Economic Research, Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/MEHOINUSTXA646N

Herzog, M., Texas Monthly Staff. (2013, February 5). *Rick Perry to Californians: Come Check Out Texas.* Texas Monthly.  https://www.texasmonthly.com/news-politics/rick-perry-to-californians-come-check-out-texas/

Internal Revenue Service. (2023, April 27). *SOI Tax Stats - Migration Data.* Tax Statistics. https://www.irs.gov/statistics/soi-tax-stats-migration-data 

National Weather Service. (2022, September). *Hurricane Katrina - August 2005.* National Oceanic and Atmospheric Administration. https://www.weather.gov/mob/katrina

O'Connor, P. (2023, September). *IRS Migration Data - 1992 to 2020.* kaggle. https://www.kaggle.com/datasets/wumanandpat/irs-migration-data-1992-to-2020




